---
title: "Sections 1.3 and 1.4: Measures of Location and Variability"
subtitle: "Mathematical Statistics"
author: "MATH 365"
format:
  revealjs:
    chalkboard: true
    theme: [default, custom.scss]
    toc: false
    toc-depth: 1
    reveal-options:
      slideNumber: true
    html-math-method: mathjax
    incremental: false
    transition: fade
    preview-links: auto
    notes: true
title-slide-attributes:
  data-background-color: "#EA6820"
from: markdown+emoji
execute:
  echo: true
  cache: false
editor:
  markdown:
    wrap: 72
---

### Warm-Up: What Value of $c$ Minimizes Squared Deviations?

We will investigate:\
$$
S(c) = \sum_{i=1}^3 (x_i - c)^2 \quad \text{for } x_1 = 4,\ x_2 = 7,\ x_3 = 10
$$

------------------------------------------------------------------------

### Warm-Up

1.  Choose a value for $c$ (e.g., $5,\ 6,\ 7,\ 8$). Compute
    $S(c) = (x_1 - c)^2 + (x_2 - c)^2 + (x_3 - c)^2$. [What do you
    notice?]{.fragment}

2.  [Try to guess the value of $c$ that minimizes $S(c)$. Why does that
    value make sense?]{.fragment}

3.  [Optional extension: Expand and simplify $S(c)$ algebraically. Can
    you take its derivative?]{.fragment}

4.  [What do you think this says about the mean as a "best" measure of
    center?]{.fragment}

------------------------------------------------------------------------

## Topics for Today

::: incremental
-   Measures of center and location
-   Measures of spread
-   Visualizations for comparing groups
:::

## Data Set: Chesapeake Total Nitrogen Load Data {.smaller}

::: incremental
-   **Environmental Motivation**: The Clean Water Act requires U.S.
    waters to meet pollution reduction goals to ensure they remain
    *fishable and swimmable*.
-   **Measurement Focus**: Daily **Total Nitrogen (TN)** loads (kg
    N/day) at a site in the **Chesapeake Bay** watershed.
-   **Purpose**: The study examines how commonly used methods for
    estimating pollutant loads in watersheds can produce misleading
    results, emphasizing the need for sound statistical practices.
-   **Source**: Data are drawn from *“Spurious Correlation in the USEPA
    Rating Curve Method for Estimating Pollutant Loads”*\
    (*J. of Environ. Engr.*, 2008: 610–618).
:::

::: notes
We'll use this sample to examine graphical summaries and discuss
foundational assumptions for inference.
:::

## Importing the Data

Use R to read the data into a dataset called **eg1p20**.

```{r}
eg1p20 <- read.csv("http://people.kzoo.edu/enordmoe/math365/data/exp01-20.csv")
```

## Data Analysis Tasks

1.  Plot and describe the distribution of the data
2.  Obtain measures of center
3.  Obtain measures of variability
4.  Identify and plot potential outliers

::: notes
Use R to work through these with reference to statistics of interest.\
\* Create a dotplot and histogram \* Summary statistics: measures of
center and spread (including trimmed mean) \* Five number summary \*
Boxplot
:::

# Measures of Center {background-color="#FAD9C7"}

## Mean

**Our Data**: A sample of $n$ observations $x_1, x_2,\ldots, x_n$ on the
random variable $X$.

::: {.callout-note title="Definition"}
The **mean** $\bar{x}$ of the $n$ values $x_1, x_2, \ldots, x_n$ is
given by: $$
\bar{x} =  \frac{\sum_{i=1}^n x_i}{n}.
$$
:::

## Properties of the Mean

::: incremental
-   The *center of gravity* of the distribution
    -   A histogram or dotplot would balance at $\bar{x}$
-   Deviations $x_i - \bar{x}$ sum to 0
-   Minimizes the sum of squared deviations around $\bar{x}$
-   But, the mean is **very** sensitive to outliers
:::

## Median

::: {.callout-note title="Definition"}
The **median** of the $n$ values $x_1, x_2,\ldots, x_n$ is the value
that most nearly lies in the middle with half the values smaller and
half larger.
:::

## Finding the Median

-   Arrange the data in ascending order
-   If $n$ is **odd**, median is the center value
-   If $n$ is **even**, median is the **mean** of the two center values

::: {.callout-tip icon="👉"}
**Try this:** Investigate the properties of the Mean and Median using
[Moore's
applet](https://digitalfirst.bfwpub.com/stats_applet/stats_applet_6_meanmed.html)
:::

The **median** is preferred over the mean when a **robust** measure of
center is needed.

## A Compromise Measure: Trimmed Mean

::: {.callout-note title="Definition"}
The **trimmed mean** removes a fixed proportion of the smallest and
largest values from the data, then computes the mean of the remaining
values.

-   For example, the 10% trimmed mean discards the lowest 10% and
    highest 10% of observations before averaging.
:::

In R:

``` r
mean(x, trim = 0.10)
```

# Measures of Spread {background-color="#FAD9C7"}

## The Range

::: {.callout-note title="Definition"}
The **range** is a single number:\
$$\text{Range} = \max(x_i) - \min(x_i)$$
:::

-   Measures the total spread of the data
-   **Highly sensitive** to extreme values (not robust)
-   Not very useful in general but still often used!

## Standard Deviation

::: {.callout-note title="Definition"}
The **standard deviation** $s$ of the $n$ values $x_1, x_2, \ldots, x_n$
is given by: $$
s = \sqrt{\frac{1}{n - 1} \sum_{i=1}^n (x_i - \bar{x})^2}
$$

The **sample variance** is $s^2$.
:::

## Properties of the Standard Deviation $s$

-   Measures *typical distance* from the mean
-   Like $\bar{x}$, not resistant to outliers
-   SD is 0 if all values are equal
-   Units match those of the original data
-   Denominator is $n-1$ rather than $n$ to obtain a better estimate of
    the population value $\sigma^2$.
-   The standard deviation is most useful when working with Normal data.

## Hinges and Fourths

-   Order the $n$ observations from smallest to largest.
-   Split the data into lower and upper halves.
-   If $n$ is odd, include the median $\tilde{x}$ in **both** halves.
-   The **lower fourth** or **lower hinge** ($h_l$) is the median of the
    lower half.
-   The **upper fourth** or **upper hinge** ($h_u$) is the median of the
    upper half.

## Fourth Spread {.smaller}

::: {.callout-note title="Definition"}
The fourth spread $f_s$ is John Tukey's robust measure of variability
defined: $$
    f_s = h_u - h_l
$$
:::

::: fragrment
The fourth spread $f_s$ is:

-   The **range of the middle half** of the data.

-   Unambiguous and easy to calculate by hand for small datasets

-   Very similar to the Interquartile Range
:::

## The Standard Boxplot

-   Displays the **five-number summary**:
    -   $x_{\min}$, $h_l$, $\tilde{x}$, $h_u$, $x_{\max}$
-   The box spans $f_s$, with:
    -   Edges at $h_l$ and $h_u$
    -   Vertical line at the median $\tilde{x}$
    -   Whiskers extending to min and max

## The Modified Boxplot

-   Whiskers extend only to the **most extreme non-outliers**
-   **Outliers** are plotted separately
-   Some software (but not base R) uses different symbols for:
    -   **Mild outliers** (between $1.5f_s$ and $3f_s$ from the nearest
        hinge)
    -   **Extreme outliers** (more than $3f_s$ away from the nearest
        hinge)

# More Options for Subgroup Analysis {background-color="#FAD9C7"}

## Options for Comparing Statistical Summaries by Subgroup in Base R {.smaller}

#### Summarizing `mpg` by `cyl` in `mtcars`

-   Base R offers several concise tools to compute and compare summary
    statistics across groups

```{r, eval = FALSE}
# Means by cylinder group
tapply(mtcars$mpg, mtcars$cyl, mean)

# Five-number summaries by group
by(mtcars$mpg, mtcars$cyl, summary)

# Mean and SD per group as a data frame
aggregate(mpg ~ cyl, mtcars, function(x) c(mean = mean(x), sd = sd(x)))
```

::: fragment
-   :white_check_mark: `tapply()` → compact results, one stat at a time,
    useful in scripting
-   :white_check_mark: `by()` → readable summaries for each group
-   :white_check_mark: `aggregate()` → clean tabular format, good for
    displaying multiple summaries together
:::

## Side-by-side histograms: MPG by Cylinder {.smaller}

::: panel-tabset
### Plot

```{r, echo = FALSE}
# Save original graphical parameters
op <- par(no.readonly = TRUE)

# Layout: 3 rows, 1 column
par(mfrow = c(3, 1), mar = c(4, 4, 2, 1))

# Common x-axis limits
xlims <- range(mtcars$mpg)

# Histograms for each cylinder group
hist(mtcars$mpg[mtcars$cyl == 4], main = "4 cyl", xlab = "MPG", 
     xlim = xlims, col = "skyblue")
hist(mtcars$mpg[mtcars$cyl == 6], main = "6 cyl", xlab = "MPG", 
     xlim = xlims, col = "forestgreen")
hist(mtcars$mpg[mtcars$cyl == 8], main = "8 cyl", xlab = "MPG", 
     xlim = xlims, col = "tomato")

# Restore original settings
par(op)
```

### Code

``` r
# Save original graphical parameters
op <- par(no.readonly = TRUE)

# Layout: 3 rows, 1 column
par(mfrow = c(3, 1), mar = c(4, 4, 2, 1))

# Common x-axis limits
xlims <- range(mtcars$mpg)

# Histograms for each cylinder group
hist(mtcars$mpg[mtcars$cyl == 4], main = "4 cyl", xlab = "MPG", 
     xlim = xlims, col = "skyblue")
hist(mtcars$mpg[mtcars$cyl == 6], main = "6 cyl", xlab = "MPG", 
     xlim = xlims, col = "forestgreen")
hist(mtcars$mpg[mtcars$cyl == 8], main = "8 cyl", xlab = "MPG", 
     xlim = xlims, col = "tomato")

# Restore original settings
par(op)
```
:::

# Practice Exercises {background-color="#E3F2FD"}

Analyze the built-in **iris** data set.

-   Browse help using `?iris`. (Find pictures of irises:)

-   Investigate petal width using methods from class including data
    display and summary statistics.

-   Explore the relationship between petal width and species.

## Summary: Measures of Location and Spread {.smaller background-color="#E8F5E9"}

-   The **mean** minimizes squared deviations but is **sensitive to
    outliers**
-   The **median** is a **robust** alternative, especially for skewed
    data
-   The **trimmed mean** offers a compromise: reduces outlier impact
    while retaining more data
-   The **standard deviation** measures typical variability but is not
    resistant
-   The **fourth spread** is a robust measure of spread, similar to the
    IQR
-   **Boxplots** provide visual summaries:
    -   Standard: min, hinges, median, max
    -   Modified: highlights mild and extreme outliers
-   Base R tools like `summary()`, `fivenum()`, and `boxplot()` support
    both numeric and visual analysis
-   For grouped data: use `tapply()`, `by()`, or `aggregate()` to
    compare summaries across subgroups
